Step by Step: Using Step Selection Functions to Simulate Dispersal and Assess Landscape Connectivity

### 2021-06-24 Chapter_1/04_Manuscript/Manuscript.tex
various updates

%------------------------------------------------------------------------------
%	Abstract
%------------------------------------------------------------------------------
\newpage
\begin{abstract}
Dispersal is an important process that allows species to avoid inbreeding, to
colonize new habitats and to reinforce non-viable subpopulations. Successful
dispersal thus represents a crucial pre-requisite for long-term species
persistence in wild animal populations. However, the ability to disperse is
contingent a sufficient degree of landscape connectivity, which is why the
estimation of connectivity and identification and preservation of dispersal
corridors has become a task of extraordinary importance for conservation
authorities worldwide.

Over the past two decades, ecologists have primarily relied on analytical tools
such as least-cost analysis and circuit theory to model and investigate
landscape connectivity. Despite their usefulness for a diverse suite of
ecological applications, both methods make several restricting assumptions that
limit their suitability in reality. To address these shortcomings, dispersal
simulations from individual-based movement models have been proposed and
applied. Yet, due to the almost infinite number of non-trivial decisions a
modeler faces when parametrizing such models, a unified and objective framework
is missing.

Recent innovations in movement ecology have brought forward novel opportunities
to study animal dispersal and estimate landscape connectivity. In particular,
the rich suite of resource selection functions, namely point-, step-, and
path-selection functions, have undergone substantial improvements over the past
years. Most notably, step-selection functions have been generalized to
\textit{integrated} step selection functions, which essentially represent fully
mechanistic movement models based on which an individual's movement could be
simulated. While such models have been applied to study \textit{steady-state}
utilization distribution resident animals, a similar approach may be useful for
investigating \textit{transient} movement behavior and study landscape
connectivity.

Here, we showcase the use of integrated step selection functions as a simple
individual-based and spatially explicit model to simulate dispersal of the
endangered African wild dog across the world's largest transboundary
conservation area, the Kavango-Zambezi Transfrontier Conservation Area
(KAZA-TFCA). For this, we utilize data collected on 16 dispersing wild dog
coalitions in combination with relevant habitat covariates. We analyse the data
using integrated step selection functions and parametrize a fully mechanistic
model describing how dispersing wild dogs move through the landscape. Based on
this model, we simulate 80'000 dispersers moving across the extent of the
KAZA-TFCA. We then generate a set of maps, each focused on a different aspect of
connectivity within the KAZA-TFCA. Finally, we discuss the benefits and pitfalls
of such a simulation-based approach and highlight potential improvements to be
made in the future.

@@

% Importance of Dispersal & Connectivity
\subsection{Importance of Dispersal \& Connectivity (90\%)}
Dispersal is defined as the movement of individuals away from their natal
location to the site of first reproduction \cite{Howard.1960}. It is a vital
process governing the social structure of wild animal populations that are
distributed in space \citep{Hanski.1998, Clobert.2012} and may strongly affect
population dynamics at different spatial and social scales \citep{Hanski.1999a,
Clobert.2012}. Dispersal allows species to avoid inbreeding and maintain genetic
diversity \citep{Perrin.1999, Perrin.2000, Frankham.2002, Leigh.2012,
Baguette.2013}, to rescue small, non-viable populations \citep{Brown.1977}, and
to promote the colonization of unoccupied habitats \citep{Hanski.1999b,
MacArthur.2001}. However, successful dispersal requires a sufficient degree of
landscape connectivity \citep{Fahrig.2003, Clobert.2012}, which is why the
identification and protection of major dispersal corridors has become a
fundamental task in conservation science \citep{Nathan.2008, Doerr.2011,
Rudnick.2012}. Importantly, the ability to pinpoint relevant dispersal hotspots
requires information on movement behavior during dispersal and knowledge about
factors that limit dispersal and therefore connectivity \citep{Baguette.2013,
Vasudev.2015}.

@@

% Resource Selection & Connectivity
\subsection{Resource Selection \& Connectivity (90\%)}
\textit{Resource selection functions} \citep{Boyce.2002} and derived methods
such as \textit{step selection functions} \citep{Fortin.2005} and \textit{path
selection functions} \citep{Cushman.2010} have proven particularly useful for
studying animal movement \citep{Fieberg.2020} and modeling connectivity
\citep{Diniz.2020}. These methods allow estimating habitat preferences of the
focal species by comparing covariates at locations visited by the animal to the
same covariates at locations available to, but not visited by the animal
\citep{Boyce.2002, Fortin.2005, Cushman.2010, Thurfjell.2014}. The so estimated
preferences can then be used to predict a permeability surface, indicating the
expected ease at which an animal can traverse a given area \citep{Spear.2010,
Zeller.2012, Etherington.2016}. Ultimately, the permeability surface serves as
input to a connectivity model that is used to reveal movement corridors
\citep{Diniz.2020}. In this regard, two of the most prominent connectivity
models are least-cost path analysis (LCP analysis; \citealp{Adriaensen.2003})
and circuit theory (CT \citealp{McRae.2006, McRae.2008}), both graph-based
methods that estimate conductance of the landscape. Despite their intuitive
nature and ease of use, both methods make rigorous assumptions about animal
movement that are often not fulfilled in reality \citep{Diniz.2020}.

% Issues with Least-Cost Paths & Circuit Theory
\subsection{Issues with Least-Cost Paths (90\%)}
In LCP analysis, for instance, a least costly path always exists, even if
associated movement costs are unreasonably high and will never be incurred by a
dispersing individual. The method also presumes that animals have an infinite
perceptual range and a preconceived end-point in mind, such that they choose a
cost-minimizing route accordingly. These assumptions may be reasonable for
migrating animals, yet they are unlikely to hold true for dispersers, which
typically move over long distances into unknown territory \citep{Koen.2014,
Abrahms.2017, Cozzi.2020}. Another issue of LCPs analysis concerns the fact that
least-costly routes, by their very nature, are only one pixel wide. This implies
that the width of inferred movement routes depends on the resolution of chosen
covariate layers and may not be biologically meaningful \citep{Diniz.2020}.
Although some of these deficiencies can be addressed using less stringent
versions of the LCP algorithm (e.g. least-cost \textit{corridors}
\citep{Pinto.2009}, \textit{thresholded} least-cost paths \citep{Landguth.2012},
and \textit{randomized} least-cost paths \citep{Panzacchi.2016,
VanMoorter.2021}), a certain degree of arbitrariness remains.

\subsection{Issues with Circuit Theory (90\%)}
CT entails similarly unreasonable restrictions that are hardly ever met. Because
CT only allows movements from a source cell to its 4 or 8 adjacent cells, it
implicitly posits a perceptual range of a single pixel. Assuming that covariate
layers are usually resolve between 30 m and 1 km, this hardly ever renders the
true capability of animals to perceive the environment. Moreover, CT builds on
the assumption of a complete random walk \citep{Diniz.2020}, implying that
directional biases cannot be rendered. Nevertheless, directionality is a
commonly observed characteristic in animal movement \citep{Bovet.1991,
Schultz.2001}, especially in dispersing individuals \citep{Cozzi.2020,
Hofmann.2021}.

\subsection{Issues of Both Methods}
Finally, neither LCP analysis nor CT are capable of rendering the temporal
dimension of dispersal. Statements about the expected duration required to
traverse a certain corridor are therefore impossible. Likewise, because movement
is not modeled explicitly, neither of the methods allows to render interactions
between movement and habitat preferences of the focal species. Connectivity
therefore merely arises in result to the landscape structure, which is usually
referred to as structural connectivity. While structural connectivity yields
insights in the \textit{potential} of the landscape to be traversed, it does not
enable to quantify the \textit{actual} gene flow through the area. Consequently,
a functional on connectivity, which also renders the behavioral response of the
animal with respect to prevailing habitat conditions, is often more desirable
\citep{Tischendorf.2000, Baguette.2013}.

% What about IBMMs?
\subsection{What about IBMMs? (90\%)}
To address the issues inherent to LCPs and CT, individual-based movement models
(IBMMs) have been proposed and applied \citep{Diniz.2020}. In these models,
dispersal is simulated spatially explicitly, based on movement rules that
determine how individuals move over and interact with the prevailing landscape
\citep{Gustafson.1996, Gardner.2004, Graf.2007, KramerSchadt.2004, Revilla.2004,
Revilla.2008, Kanagaraj.2013, Clark.2015, Allen.2016, Hauenstein.2019,
Zeller.2020, Vasudev.2021}. Using the simulated trajectories, one can calculate
a set of connectivity metrics, such as inter-patch connectivity and traversal
frequency to reveal major dispersal corridors \citep{Kanagaraj.2013,
BastilleRousseau.2018, Hauenstein.2019, Zeller.2020}. However, while IBMMS can
be employed to overcome any of the shortcomings intrinsic to LCPs and CT and
provide a more functional view on connectivity, they can be challenging to fit
and require vast amounts of data collected during dispersal \citep{Diniz.2020}.
Moreover, most individual-based models are heavily influenced by subjective
decisions, rendering replication and validation of results difficult.

% Step Selection Analysis
\subsection{Step Selection Analysis (90\%)}
Here, we investigate the usefulness of integrated step selection functions
(ISSFs, \citealp{Avgar.2016}), as a relatively simple but powerful IBMM based on
which dispersal can be simulated. While regular SSFs were intended to learn
about relative habitat preferences of the focal species \citep{Fortin.2005,
Thurfjell.2014, Avgar.2017}, the method has been generalized and now enables to
jointly study habitat and movement preferences, as well as potential
interactions between movement and habitat preferences \citep{Avgar.2016,
Signer.2017, Fieberg.2020}. ISSFs therefore provide a relatively simple method
to model complex movement behavior where movement results from two intertwined
behavioral kernels (e.g. \citealp{Prokopenko.2017, Munden.2020}). Importantly, a
parametrized ISSF model can be viewed as a fully mechanistic movement model
based on which individual movement trajectories can be simulated
\citep{Avgar.2016, Signer.2017}. In fact, \cite{Signer.2017} used ISSF to
simulate steady state utilization distributions of resident animals. However,
the degree to which such simulations are helpful in detecting movement corridors
and modeling landscape connectivity is unknown.

% Study Species & Study Area
\subsection{Study Species \& Study Area (90\%)}
One of the species for which long-term viability relies on sufficient landscape
connectivity is the endangered African wild dog \textit{Lycon pictus}. While
once present across entire sub-Saharan Africa, wild dogs have disappeared from a
vast majority of their historic range due to persecution by humans, habitat
fragmentation and destruction, and deadly diseases \citep{Woodroffe.2012}. As of
today, only 6'000 free-ranging individuals remain in small and spatially
scattered subpopulations \citep{Woodroffe.2012}. Within those subpopulations,
wild dogs form cohesive packs comprising 8 to 12 adults and their offspring
\cite{McNutt.1995}. After reaching sexual maturity, male and female offspring
form same-sex coalitions and disperse from their natal pack \citep{McNutt.1996,
Behr.2020}. New packs are formed when dispersing coalitions join unrelated
opposite-sex dispersing coalitions \citep{McNutt.1996}. Dispersing wild dogs can
cover several hundred kilometers across a variety of landscapes
\citep{DaviesMostert.2012, Masenga.2016, Cozzi.2020, Hofmann.2021}. One of the
few strongholds for this species lies near the Moremi Game Reserve in northern
Botswana, which is part of the world's largest transboundary protected area,
namely the Kavango-Zambezi Transfrontier Conservation Area (KAZA-TFCA). This
area has originally been intended to facilitate migration of elephants, but is
expected to benefit a multitude of other species \citep{Elliot.2014,
Brennan.2020, Hofmann.2021}.

% Previous Paper
\subsection{Previous Paper (90\%)}
In a previous study, we assessed landscape connectivity for dispersing African
wild dogs within the KAZA-TFCA using a least-cost methods \citep{Hofmann.2021}.
Specifically, we fitted a basic habitat selection model using which we predicted
a permeability surface and computed least-cost paths and corridors. We now
expand this model and use ISSFs to develop a more detailed movement model of
dispersing wild dogs (\Cref{GraphicalAbstract}). We then employ the model to
simulate dispersers moving across the KAZA-TFCA. Based on simulations, we
generate three distinc connectivity maps, all sheding light onto a different
aspect of connectivity. With this work we hope to show that a simulation-based
approach yields several benefits over traditional connectivity modeling
techniques such as least-cost analyses and circuit theory. Most importantly,
simulations provide a more generic view on how connectivity emerges and to which
degree connectivity depends on the dispersal duration. In addtion, by generating
proper dispersal trajectories, network theory can be applied to calculate
network metrics that are pertinent to connectivity analysis.

@@

\section{Methods}
\subsection{Study Area (90\%)}
The study area was defined by a bounding box centered at -17\degree 13'9''S,
23\degree 56'4''E (\Cref{StudyArea}a) stretching over 1.3 Mio.
km\textsuperscript{2} and ecompassing the entire KAZA-TFCA (\Cref{StudyArea}b).
The KAZA-TFCA represents the world's largest transboundary conservation area and
comprises parts of Angola, Botswana, Namibia, Zimbabwe, and Zambia. It covers a
total of 520'000 km\textsuperscript{2} and hosts a diverse landscape, ranging
from savanna to grassland and from dry to moist woodland habitats. In its center
lies the Okavango Delta, a dominant hydrogeographical feature and the world's
largest flood-pulsing inland delta. The wet season within the KAZA-TFCA lasts
from November to March and is out of phase with the flood in the Okavango Delta,
which peaks between July and August \citep{McNutt.1996, Wolski.2017}. Although
large portions within the KAZA-TFCA are designated national parks or other
protected areas, considerable human influence remains due to roads, agricultural
sites and settlements and villages that are distributed across the KAZA-TFCA's
landscape.

@@

16 collared individuals eventually dispersed, each in a separate same-sex
dispersal coalition (7 female and 9 male coalitions). During dispersal, collars
were programmed to record a GPS fix every 4 hours. Collected relocations were
regularly transmitted over the Iridium satellite system, which allowed remote
tracking of individuals, even if they left the main study area and crossed
international borders. Because behavior during dispersal is more pertinent for
assessing landscape connectivity \citep{Elliot.2014, Abrahms.2017}, we discarded
all data that was collected during residency and only retained GPS data recorded
during dispersal. In some instances, exact dispersal dates were known from field
observations. Where this was not the case, determined dispersal phases using the
net-squared displacement metric. Net squared displacement measures the squared
Euclidean distance of a GPS relocation to a reference point \citep{Borger.2012},
which in our case was set to the center of each individual's natal home range.
As such, dispersal was deemed to have started when an individual left its natal
home range and ended once individuals became sedentary again. As previous
research revealed similar behavior of females and males during dispersal
\citep{Woodroffe.2019, Cozzi.2020}, we did not distinguish between sexes. After
collection, we converted collected GPS coordinates (n = 4'169) to steps, where
each step represented the straight-line distance traveled by and individual
between two consecutive GPS relocations \citep{Turchin.1998}. To ensure a
regular sampling interval, we removed fixes that were not successfully collected
on the 4-hourly schedule (\( \pm \) 15 minutes).

@@

\subsection{Movement Model (80\%)}
We used ISSFs to parametrize a mechanistic movement model for dispersing wild
dogs \citep{Avgar.2016}. To conduct ISSF analysis, we paired each realized step
with 24 random steps. An observed and its 24 random steps formed a stratum and
received a unique identifier. We generated random steps by sampling random
turning angles from a uniform distribution (\(-\pi, +\pi\)) and step lengths
from a gamma distirbution that was fitted to realized steps (scale = 6'308,
shape = 0.37). Along each step, we extracted and averaged spatial covariates
using the {\tt velox} package \citep{Hunziker.2021}. We also calculated the
movement metrics \textsf{sl}, \textsf{log(sl)}, and \textsf{cos(ta)} for each
observed and random step. To facilitate model convergence, we standardized all
continuous covariates to a mean of zero and a standard deviation of one. Since
correlation among covariates was low (\(|r| > 0.6\); \citealp{Latham.2011}), we
retained all of them for modeling.

To contrast realized steps (scored 1) and random steps (scored 0), we assumed
that animals assigned a selection score \(w(x)\) of the exponential form to each
step \citep{Fortin.2005}. The selection score \(w(x)\) of each step depended on
its associated covariates (\(x_1, x_2, ..., x_n\)) and on the animal's
preferences (i.e. relative selection strengths; \citealp{Avgar.2017}) towards
these covariates (\(\beta_1, \beta_2, ..., \beta_n\)):

@@

To estimate the preferences of interest, we ran conditional logistic regression
analysis in the r-package {\tt glmmTMB}. To handle multiple individuals, we
applied the mixed effects technnique developed by \citep{Muff.2020}, which
allows to effectively model random slopes. Thus, we treated animal IDs as random
effect and modeled random slopes for each covariate. We fixed the random
intercept variance at an arbitrary high value of 10\textsuperscript{6} to make
use of the  ``poission''-trick \citep{Muff.2020}.

The formula for the movement model was based on the habitat selection model for
dispersing wild dogs presented in \cite{Hofmann.2021}. In the original model, no
interactions between habitat covariates (\textsf{Water,
DistanceToWater\textsuperscript{0.5}, Woodland, Shrubs/Grassland, Human
Influence}) and movement covariates (\textsf{sl, log(sl), cos(ta)}) were
considered. Here, we slightly expanded this base model and proposed interactions
between all movement and habitat covariates. More specifically, we started with
the base model and incrementally increased model complexity by adding all
possible two-way interactions between habitat covariates and movement
covariates. For instance, for the covariate \textsf{Water}, we proposed the
interactions \textsf{Water:log(sl)}, \textsf{Water:log(sl)}, and
\textsf{Water:cos(ta)}. Besides those combinations, we also proposed the
interactions \textsf{sl:cos(ta)} and \textsf{log(sl):cos(ta)} to account for a
correlation between turning angles and step lengths, as well as the interactions
\textsf{sl:LowActivity} and \textsf{log(sl):LowActivity} to account for the fact
that step lengths may differ due to wild dogs' diurnal activity pattern. To
compare competing models and assess the most parsimonious movement model, we ran
stepwise forward model selection based on Akaike's Information Criterion (AIC,
\citealp{Burnham.2002}).

>>>>

%------------------------------------------------------------------------------
%	Abstract
%------------------------------------------------------------------------------
\newpage
\begin{abstract}
Dispersal is an important process that allows species to avoid inbreeding,
colonize new habitats and reinforce non-viable subpopulations. Successful
dispersal thus represents a crucial pre-requisite for long-term species
persistence in wild animal populations. However, the ability to disperse is
contingent a sufficient degree of landscape connectivity, which is why the
estimation of connectivity and preservation of dispersal corridors has become a
task of extraordinary importance for conservation authorities worldwide.

Over the past two decades, ecologists have primarily relied on analytical tools
such as least-cost analysis and circuit theory to model and investigate
landscape connectivity. Despite their usefulness for a diverse suite of
ecological applications, both methods make several restricting assumptions that
limit their suitability in reality. Individual-based dispersal simulations have
been proposed to address these shortcomings, yet due to the sheer amount of
non-trivial modeling decisions decisions required, a unified and objective
framework to simulate dispersal is missing.

Recent innovations in movement ecology have brought forward novel opportunities
to study animal dispersal and estimate landscape connectivity. In particular,
the rich suite of resource selection functions, namely point-, step-, and
path-selection functions, have undergone substantial improvements over the past
years. Most notably, step-selection functions have been generalized to
\textit{integrated} step selection functions, which essentially represent fully
mechanistic movement models based on which an individual's movement could be
simulated. While such models have been applied to study \textit{steady-state}
utilization distribution resident animals, a similar approach may be useful to
investigate \textit{transient} movement behavior and landscape connectivity.

Here, we showcase the use of integrated step selection functions as a simple
individual-based and spatially explicit model to simulate dispersal of the
endangered African wild dog across the world's largest transboundary
conservation area, the Kavango-Zambezi Transfrontier Conservation Area
(KAZA-TFCA). For this, we utilize data collected on 16 dispersing wild dog
coalitions in combination with relevant habitat covariates. We analyse the data
using integrated step selection functions, thereby parametrizing a fully
mechanistic movement model describing how dispersing wild dogs move through the
landscape. Based on this model, we simulate 80'000 dispersers moving across the
extent of the KAZA-TFCA. and generate a set of maps, each focused on a different
aspect of connectivity. Finally, we discuss the benefits and pitfalls of such a
simulation-based approach and highlight potential improvements to be made in the
future.

@@

% Importance of Dispersal & Connectivity
\subsection{Importance of Dispersal \& Connectivity (90\%)}
Dispersal is defined as the movement of individuals away from their natal
location to the site of first reproduction \cite{Howard.1960}. It is a vital
process governing the social structure of wild animal populations that are
distributed in space \citep{Hanski.1998, Clobert.2012} and may strongly affect
population dynamics at different spatial and social scales \citep{Hanski.1999a,
Clobert.2012}. Dispersal allows species to maintain genetic diversity
\citep{Perrin.1999, Perrin.2000, Frankham.2002, Leigh.2012, Baguette.2013}, to
rescue small, non-viable populations \citep{Brown.1977}, and to promote the
colonization or recolonization of unoccupied habitats \citep{Hanski.1999b,
MacArthur.2001}. However, successful dispersal requires a sufficient degree of
landscape connectivity \citep{Fahrig.2003, Clobert.2012}, which is why the
identification and protection of major dispersal corridors has become a
fundamental task in conservation science \citep{Nathan.2008, Doerr.2011,
Rudnick.2012}. The ability to pinpoint relevant dispersal hotspots requires
information on movement behavior during dispersal and knowledge about factors
that limit dispersal and therefore connectivity \citep{Baguette.2013,
Vasudev.2015}.

@@

% Resource Selection & Connectivity
\subsection{Resource Selection \& Connectivity (90\%)}
\textit{Resource selection functions} \citep{Boyce.2002} and derived methods
such as \textit{step selection functions} \citep{Fortin.2005} and \textit{path
selection functions} \citep{Cushman.2010} have proven particularly useful for
studying animal movement \citep{Fieberg.2020} and modeling connectivity
\citep{Diniz.2020}. These methods allow estimating habitat preferences of the
focal species by comparing covariates at locations visited by the animal to the
same covariates at locations available to, but not visited by the animal
\citep{Boyce.2002, Fortin.2005, Cushman.2010, Thurfjell.2014}. The so estimated
preferences can then be used to predict a permeability surface, indicating the
expected ease at which an animal can traverse a given area \citep{Spear.2010,
Zeller.2012, Etherington.2016}. Ultimately, the permeability surface serves as
input to a connectivity model that is used to reveal movement corridors
\citep{Diniz.2020}. Two of the most prominent connectivity models are least-cost
path analysis (LCP analysis; \citealp{Adriaensen.2003}) and circuit theory (CT;
\citealp{McRae.2006, McRae.2008}), both graph-based methods that estimate
conductance of the landscape to infer likely movement corridors. Despite their
intuitive nature and ease of use, both methods make rigorous assumptions about
animal movement that are often not fulfilled in reality \citep{Diniz.2020}.

% Issues with Least-Cost Paths & Circuit Theory
\subsection{Issues with Least-Cost Paths (90\%)}
In LCP analysis, for instance, a least costly path always exists, even if
associated movement costs are unreasonably high and will never be incurred by a
dispersing individual. The method also presumes that animals have an infinite
perceptual range and a preconceived end-point in mind, such that they choose a
cost-minimizing route accordingly. These assumptions may be fulfilled by
migrating animals that typically move between a discrete set of habitats through
familiar landscapes. Dispersers, on the other hand, usually move over long
distances into unknown territory and are therefore less likely to be aware of
associated movement costs \citep{Koen.2014, Abrahms.2017, Cozzi.2020}. Another
issue of LCPs analysis concerns the fact that least-costly routes, by their very
nature, are only one pixel wide \citep{Pinto.2009}. This neglects the fact that
alternative routes with similar costs may exist and implies that the width of
inferred movement routes depends on the resolution of chosen covariate layers
and may not be biologically meaningful \citep{Diniz.2020}. Although some of
these deficiencies can be addressed using less stringent versions of the LCP
algorithm (e.g. least-cost \textit{corridors} \citep{Pinto.2009},
\textit{thresholded} least-cost paths \citep{Landguth.2012}, and
\textit{randomized} least-cost paths \citep{Panzacchi.2016, VanMoorter.2021}), a
certain degree of arbitrariness remains.

\subsection{Issues with Circuit Theory (90\%)}
CT entails similarly unreasonable restrictions that are hardly ever met. For
example, because CT only allows movements from a source cell to its 4 or 8
adjacent cells, it implicitly posits that animals exhibit a perceptual range of
a single pixel. Given that covariate layers are usually resolved with a pixel
size between 30 m x 30 m and 1 km x 1 km, this hardly ever renders the true
capability of animals to perceive the environment. Moreover, CT is built around
the assumption of a complete random walk \citep{Diniz.2020}, entailing that
directional biases cannot be rendered. Nevertheless, directionality is a common
characteristic in animal movement \citep{Bovet.1991, Schultz.2001}, especially
in dispersing individuals \citep{Cozzi.2020, Hofmann.2021}.

\subsection{Issues of Both Methods}
Finally, neither LCP analysis nor CT are capable of rendering the temporal
dimension of dispersal \citep{Diniz.2020}. Statements about the expected
duration required to traverse a certain corridor are therefore impossible.
Likewise, because movement is not modeled explicitly, interactions between
movement and habitat preferences of the focal species cannot be rendered.
Connectivity therefore merely arises in result to the landscape structure, which
is usually referred to as structural connectivity. While structural connectivity
yields insights in the \textit{potential} of the landscape to be traversed, it
does not enable to quantify the \textit{actual} gene flow through the area.
Consequently, a functional view on connectivity, which also renders the
behavioral response of the animal with respect to prevailing habitat conditions,
is often more desirable \citep{Tischendorf.2000, Baguette.2013}.

% What about IBMMs?
\subsection{What about IBMMs? (90\%)}
To address the issues inherent to LCPs and CT, individual-based movement models
(IBMMs) have been proposed and applied \citep{Diniz.2020}. In these models,
dispersal trajectories are simulated spatially explicitly, based on movement
rules that determine how individuals move over and interact with the prevailing
landscape \citep{Gustafson.1996, Gardner.2004, Graf.2007, KramerSchadt.2004,
Revilla.2004, Revilla.2008, Kanagaraj.2013, Clark.2015, Allen.2016,
Hauenstein.2019, Zeller.2020, Vasudev.2021}. Using the simulated trajectories,
one can calculate a set of connectivity metrics, such as inter-patch
connectivity and traversal frequency, to reveal major dispersal corridors
\citep{Kanagaraj.2013, BastilleRousseau.2018, Hauenstein.2019, Zeller.2020}.
However, while IBMMS can be employed to overcome any of the shortcomings
intrinsic to LCPs and CT, they are subject to a vast amount of subjective,
non-trivial modeling decisions. Moreover, they can be challenging to fit and
require vast amounts of movement data, ideally collected during dispersal
\citep{Diniz.2020}. Consequently, alternative methods that require fewer
modeling decisions and are straight forward to apply are desirable.

% Step Selection Analysis
\subsection{Step Selection Analysis (90\%)}
Here, we investigate the usefulness of integrated step selection functions
(ISSFs, \citealp{Avgar.2016}), as a relatively simple but powerful IBMM based on
which dispersal can be simulated. While regular SSFs were intended to learn
about relative habitat preferences of the focal species \citep{Fortin.2005,
Thurfjell.2014, Avgar.2017}, the method has recently been generalized to
\textit{integrated} SSFs and now enables to jointly study habitat and movement
preferences, as well as potential interactions between them \citep{Avgar.2016,
Signer.2017, Fieberg.2020}. ISSFs therefore provide a relatively simple means to
model complex movement behavior, where movement is viewed as the result of two
intertwined behavioral kernels (e.g. \citealp{Prokopenko.2017, Munden.2020}).
Importantly, a parametrized ISSF model can be employed as a fully mechanistic
movement model based on which individual movement trajectories can be simulated
\citep{Avgar.2016, Signer.2017}. In fact, \cite{Signer.2017} used ISSF to
simulate steady state utilization distributions of resident animals that were
moving around a point of attraction. However, the degree to which such
simulations are helpful in detecting movement corridors and modeling landscape
connectivity remains to be investigated.

% Study Species & Study Area
\subsection{Study Species \& Study Area (90\%)}
One of the species for which long-term viability relies on sufficient landscape
connectivity is the endangered African wild dog \textit{Lycon pictus}. While
once present across entire sub-Saharan Africa, wild dogs have disappeared from a
vast majority of their historic range due to persecution by humans, habitat
fragmentation and destruction, and deadly diseases \citep{Woodroffe.2012}. As of
today, only 6'000 free-ranging individuals remain in small and spatially
scattered subpopulations \citep{Woodroffe.2012}. Within those subpopulations,
wild dogs form cohesive packs comprising 8 to 12 adults and their offspring
\cite{McNutt.1995}. After reaching sexual maturity, male and female offspring
form same-sex coalitions and disperse from their natal pack in search for
potential mating partners and a suitable territory to settle \citep{McNutt.1996,
Behr.2020}. New packs are formed when dispersing coalitions join unrelated
opposite-sex dispersing coalitions \citep{McNutt.1996}. Dispersing wild dogs can
cover several hundred kilometers across a variety of landscapes
\citep{DaviesMostert.2012, Masenga.2016, Cozzi.2020, Hofmann.2021}. One of the
few strongholds for this species lies near the Moremi Game Reserve in northern
Botswana, which is part of the world's largest transboundary protected area,
namely the Kavango-Zambezi Transfrontier Conservation Area (KAZA-TFCA). This
area has originally been intended to facilitate migration of elephants, but is
expected to provide benefits to a multitude of other species \citep{Elliot.2014,
Brennan.2020, Hofmann.2021}.

% Previous Paper
\subsection{Previous Paper (90\%)}
In a previous study, we assessed landscape connectivity for dispersing African
wild dogs within the KAZA-TFCA using least-cost methods \citep{Hofmann.2021}.
Specifically, we fitted a basic habitat selection model and predicted a
permeability surface that we used to compute least-cost paths and corridors. We
now expand on this knowledge and use ISSFs to develop a more mechanistic
movement model of dispersing wild dogs (\Cref{GraphicalAbstract}). We employ the
model to simulate dispersers moving across the KAZA-TFCA and generate three
distinct connectivity maps, each shedding light onto a different aspect of
connectivity. With this work, we exemplify how ISSFs can be utilized for
dispersal simulations and we discuss several benefits of this approach over
traditional connectivity modeling techniques such as least-cost analyses and
circuit theory. Most importantly, simulations based on ISSFs provide a more
generic view on how connectivity emerges and to which degree connectivity
depends on the dispersal duration. In addition, by generating proper dispersal
trajectories, network theory can be applied to calculate network metrics that
are pertinent to connectivity analysis.

@@

\section{Methods}
\subsection{Study Area (90\%)}
The study area was defined by a bounding box centered at -17\degree 13'9''S,
23\degree 56'4''E (\Cref{StudyArea}a) stretching over 1.3 Mio.
km\textsuperscript{2} and encompassing the entire KAZA-TFCA (\Cref{StudyArea}b).
The KAZA-TFCA represents the world's largest transboundary conservation area and
comprises parts of Angola, Botswana, Namibia, Zimbabwe, and Zambia. It covers a
total of 520'000 km\textsuperscript{2} and hosts diverse landscapes, ranging
from savanna to grassland and from dry to moist woodland habitats. In its center
lies the Okavango Delta, a dominant hydrogeographical feature and the world's
largest flood-pulsing inland delta. The wet season within the KAZA-TFCA lasts
from November to March and is out of phase with the flood in the Okavango Delta,
which peaks between July and August \citep{McNutt.1996, Wolski.2017}. Although
large portions within the KAZA-TFCA are designated national parks or other
protected areas, considerable human influence remains due to roads, agricultural
sites and settlements and villages that are distributed across the KAZA-TFCA's
landscape.

@@

16 collared individuals eventually dispersed, each in a separate same-sex
dispersal coalition (7 female and 9 male coalitions). During dispersal, collars
were programmed to record a GPS fix every 4 hours, all of which were regularly
transmitted over the Iridium satellite system, thereby allowing to remotely
track individuals, even if they left the main study area and crossed
international borders. Because behavior during dispersal is more pertinent for
assessing landscape connectivity \citep{Elliot.2014, Abrahms.2017}, we discarded
all data that was collected during residency and only retained GPS data recorded
during dispersal. In some instances, exact dispersal dates were known from field
observations, whereas in other cases we determined dispersal phases using the
net-squared displacement metric. Net squared displacement measures the squared
Euclidean distance of a GPS relocation to a reference point \citep{Borger.2012},
which in our case was set to the center of each individual's natal home range.
Thus, dispersal was deemed to have started when an individual left its natal
home range and ended once individuals became sedentary again. As previous
research found no differences in behaviors of females and males during dispersal
\citep{Woodroffe.2019, Cozzi.2020}, we did not distinguish between sexes. After
collection, we converted collected GPS coordinates (n = 4'169) to steps, where
each step represented the straight-line distance traveled by an individual
between two consecutive GPS relocations \citep{Turchin.1998}. To ensure a
regular sampling interval, we removed fixes that were not successfully collected
on the 4-hourly schedule (\( \pm \) 15 minutes).

@@

\subsection{Movement Model (80\%)}
We used ISSFs to parametrize a mechanistic movement model for dispersing wild
dogs \citep{Avgar.2016}. To conduct ISSF analysis, we paired each realized step
with 24 random steps. An observed step plus its 24 random steps formed a stratum
and received a unique identifier. As suggested by \cite{Avgar.2016}, we
generated random steps by sampling random turning angles from a uniform
distribution (\(-\pi, +\pi\)) and step lengths from a gamma distribution that
was fitted to realized steps (scale = 6'308, shape = 0.37). Along each step, we
extracted and averaged spatial covariates using the {\tt velox} package
\citep{Hunziker.2021}. We also calculated the movement metrics \textsf{sl},
\textsf{log(sl)}, and \textsf{cos(ta)} for each observed and random step. To
facilitate model convergence, we standardized all continuous covariates to a
mean of zero and a standard deviation of one. Since correlation among covariates
was low (\(|r| > 0.6\); \citealp{Latham.2011}), we retained all of them for
modeling.

To contrast realized steps (scored 1) and random steps (scored 0), we assumed
that animals assigned a selection score \(w(x)\) of the exponential form to each
step \citep{Fortin.2005}. The selection score \(w(x)\) of each step thus
depended on its associated covariates (\(x_1, x_2, ..., x_n\)) and on the
animal's preferences (i.e. relative selection strengths; \citealp{Avgar.2017})
towards these covariates (\(\beta_1, \beta_2, ..., \beta_n\)):

@@

We ran conditional logistic regression analysis in the r-package {\tt glmmTMB}
to estimate preferences of interest. To handle multiple individuals, we applied
the mixed effects technique developed by \citep{Muff.2020}, which allows to
effectively model random slopes. Thus, we treated animal IDs as random effect
and modeled random slopes for each covariate. We fixed the random intercept
variance at an arbitrary high value of 10\textsuperscript{6} to make use of the
``poission''-trick \citep{Muff.2020}.

The structure of the movement model was based on the habitat selection model for
dispersing wild dogs presented in \cite{Hofmann.2021}. In the original model
(referred to as base model hereafter), no interactions between habitat
covariates (\textsf{Water, DistanceToWater\textsuperscript{0.5}, Woodland,
Shrubs/Grassland, Human Influence}) and movement covariates (\textsf{sl,
log(sl), cos(ta)}) were considered. Hence, we slightly expanded this base model
and proposed interactions between all movement and habitat covariates. More
specifically, we started with the base model and incrementally increased model
complexity by adding all possible two-way interactions between habitat
covariates and movement covariates. For instance, for the covariate
\textsf{Water}, we proposed the interactions \textsf{Water:log(sl)},
\textsf{Water:log(sl)}, and \textsf{Water:cos(ta)}. Besides those combinations,
we also proposed the interactions \textsf{sl:cos(ta)} and
\textsf{log(sl):cos(ta)} to account for a correlation between turning angles and
step lengths, as well as the interactions \textsf{sl:LowActivity} and
\textsf{log(sl):LowActivity} to account for the fact that step lengths may
differ due to wild dogs' diurnal activity pattern. To compare competing models
and assess the most parsimonious movement model, we ran stepwise forward model
selection based on Akaike's Information Criterion (AIC, \citealp{Burnham.2002}).
